\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-08 analysis/paper/paper.Rmd
fix table captions

# Code formatting

Don't use markdown, instead use the more precise latex commands:

* \proglang{Java}
* \pkg{plyr}
* \code{print("abc")}
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

hazard ratios
2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes!
[@hanley2009fitting]

@@

These packages can belong to three main categories: non-proportional hazards, regularization and competing risks. Besides these three main categories, there are extra features of interest as well. Table XXX summarizes the findings.

Many packages exist that implement semi-parametric cox models. Of those, glmnet [-@regpathcox]???, glmpath [-@park_hastie], penalized [-@l1penal], RiskRegression [-@gerds_blanche] introduce penalization in their model fitting. CoxRidge [-@perperoglou] , rstpm2 [-@clements_liu] and survival [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not. In terms of penalization, the benefit of casebase comes from its flexibility. Linear models have a wide tool set, considerably larger than those made for cox regression (*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others. The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*). Casebase permits many regularization tools developed for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in a handful of packages. Of interest are CFC [-@sharabiani_mahani_2019], flexsurv [-@flexsurv] , SmoothHazard [-@smoothHazard], Rsptm2 [-@clements_liu] and Survival [-@survival-package]. Smooth hazard  is limited to weibull distributions[-@smoothHazard], where as flexsurv and survival contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv]. Flexsurv, smooth hazard, mets and rstpm2 can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu] . Though casebase is limited to 3 distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.

Of the methods mentioned so far, only CFC,flexsurv, mets and survival contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv],[-@scheike2014estimating] ,[-@survival-package] (*how is it done for parametric models?*). The differentiating factor between these packages and casebase is that casebase handles competing risks through multiple logistic regression. 
Smooth hazard, survival and Rstpm2 can handle interval censoring, where as casebase currently can not. Interval censoring is simple feature to implement and is a potential improvement to add to casebase. (There's also multistate models, which I haven't mentioned in this paper)

Of all the packages listed in the table, CFC , Flexsurv, mets, RiskRegression, rstpm2 and survival have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package]. Since casebase is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence. This is similar to the approaches of CFC and flexsurv [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche]. rstpm2 and survival use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].

\begin{table}[]

@@

\label{tab:my-table}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.  Imputation was conducted using the default methods in the mice package in R, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual. A description of each variable can be found in table XXX1 and a breakdown of each categorical variable in table XXX2. The data was downloaded from the main author's website, and it is also available as part of the casebase package.

\begin{table}[]

@@

\label{tab:my-table}

@@

\begin{table}[]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:my-table}
\end{table}

@@

Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
legend("bottom", 
       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
       col = c("black","red","Blue"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```
# Discussion

>>>>

<!-- # Code formatting -->

<!-- Don't use markdown, instead use the more precise latex commands: -->

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->
<!-- * \code{print("abc")} -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

  1. hazard ratios
  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]

@@

These packages can belong to three main categories: non-proportional hazards, regularization and competing risks. Besides these three main categories, there are extra features of interest as well. Table \ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting. \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not. In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility. Linear models have a wide tool set, considerably larger than those made for Cox regression. <!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others. The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in a handful of packages. Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv]. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu]. Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package]. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot. Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package]. Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence. This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche]. \pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].

\begin{table}[ht]

@@

\label{tab:surv-pkgs}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.  Imputation was conducted using the default methods in the mice package in R, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the casebase package.

\begin{table}[ht]

@@

\label{tab:support1}

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:support2}
\end{table}

@@

Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
legend("bottom", 
       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
       col = c("black","red","Blue"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

# Discussion
